Why there are more species in several small patches vs few large patches:

### 2023-12-04 RunSims.Rmd
Updated manuscript

```{r PlotsNoSelectionRegular, echo=FALSE,message=FALSE,warning=FALSE,fig.align='center',eval=FALSE}

@@

# Shannon-Diversity  
#
mdl1 %>%  
  ggplot(aes(x = habitat_patch_size, y = calc_shannon_diversity_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = .5) +
  geom_point(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_shannon_diversity_ql, ymax = calc_shannon_diversity_qh, group = interaction(birds_dispersal_distance, birth_rate_birds), fill = factor(birds_dispersal_distance)), alpha = 0.2,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() + scale_fill_viridis_d() +
  facet_wrap(~ birth_rate_birds, ncol = 4) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Habitat Patch Size") + ylab("Shannon Diversity") +
  theme(
    legend.position = c(0.9,0.8)
    )

@@

mdl <- sloss_sim %>% filter(birds_behavior=="NoSelection", habitat_patch_size != 199)
names(mdl)
#
# summarize data 
#
mdl1 <- mdl %>% mutate(habitat_patch_size = habitat_patch_size+1) %>%
                         group_by(birds_dispersal_distance, habitat_patch_size,birth_rate_birds,deforestation_type) %>% summarise(across(c(birds_at_deforestation:count_birds,mean_patch,degraded_mfp,non_degraded_mfp),list(mean=mean, sd=sd, ql=~quantile(.x,probs=c((1-0.95)/2)), qh=~quantile(.x,probs=c(1-(1-0.95)/2)) ))) #  %>% mutate(deltaPop = (count_birds - birds_at_deforestation )/ birds_at_deforestation  )
names(mdl1)

@@

mdl1 %>%  filter(birth_rate_birds == 2) %>% 
  ggplot(aes(x = mean_patch_mean, y = calc_number_of_species_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = .5) +
  geom_point(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_number_of_species_ql, ymax = calc_number_of_species_qh, group = interaction(birds_dispersal_distance, deforestation_type), fill = factor(birds_dispersal_distance)), alpha = 0.3,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() + scale_fill_viridis_d() +
  facet_wrap(~ deforestation_type, ncol = 2) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Habitat Patch Size") + ylab("Number of species") +
  theme(
    legend.position = c(0.9,0.8)
    )

# Shannon-Diversity  
#
mdl1 %>%  filter(birth_rate_birds == 2) %>% 
  ggplot(aes(x = mean_patch_mean, y = calc_shannon_diversity_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = .5) +
  geom_point(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_shannon_diversity_ql, ymax = calc_shannon_diversity_qh, group = interaction(birds_dispersal_distance, deforestation_type), fill = factor(birds_dispersal_distance)), alpha = 0.3,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() + scale_fill_viridis_d() +
  facet_wrap(~ deforestation_type, ncol = 2) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Mean Habitat Patch Size") + ylab("Number of species") +
  theme(
    legend.position = c(0.9,0.8)
    )

@@

>>>>

```{r PlotsNoSelectionRegular, echo=FALSE,message=FALSE,warning=FALSE,fig.align='center',eval=FALSE}
# read data
#
sloss_sim <- readRDS("Simulations/sloss_sim_ff.rds")

@@

# Shannon-Diversity  
#
mdl1 %>%  
  ggplot(aes(x = habitat_patch_size, y = calc_shannon_diversity_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = .5) +
  geom_point(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_shannon_diversity_ql, ymax = calc_shannon_diversity_qh, group = interaction(birds_dispersal_distance, birth_rate_birds), fill = factor(birds_dispersal_distance)), alpha = 0.2,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() + scale_fill_viridis_d() +
  facet_wrap(~ birth_rate_birds, ncol = 4) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Habitat Patch Size") + ylab("Shannon Diversity") +
  theme(
    legend.position = c(0.9,0.2)
    )

@@

mdl <- sloss_sim %>% filter(birds_behavior=="NoSelection", habitat_patch_size != 199)
names(mdl)
#
# summarize data 
#
mdl1 <- mdl %>% mutate(habitat_patch_size = habitat_patch_size+1) %>%
                         group_by(birds_dispersal_distance, habitat_patch_size,birth_rate_birds,deforestation_type) %>% summarise(across(c(birds_at_deforestation:count_birds,max_patch,degraded_mfp,non_degraded_mfp),list(mean=mean, sd=sd, ql=~quantile(.x,probs=c((1-0.95)/2)), qh=~quantile(.x,probs=c(1-(1-0.95)/2)) ))) #  %>% mutate(deltaPop = (count_birds - birds_at_deforestation )/ birds_at_deforestation  )
names(mdl1)

mdl1 %>%  filter(birth_rate_birds == 2) %>% 
  ggplot(aes(x = max_patch_mean, y = calc_number_of_species_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = .5) +
  geom_point(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_number_of_species_ql, ymax = calc_number_of_species_qh, group = interaction(birds_dispersal_distance, deforestation_type), fill = factor(birds_dispersal_distance)), alpha = 0.3,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() + scale_fill_viridis_d() +
  facet_wrap(~ deforestation_type, ncol = 2) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Max Habitat Patch Size") + ylab("Number of species") +
  theme(
    legend.position = c(0.9,0.8)
    )

# Shannon-Diversity  
#

mdl1 %>%  filter(birth_rate_birds == 2) %>% 
  ggplot(aes(x = max_patch_mean, y = calc_shannon_diversity_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = .5) +
  geom_point(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_shannon_diversity_ql, ymax = calc_shannon_diversity_qh, group = interaction(birds_dispersal_distance, deforestation_type), fill = factor(birds_dispersal_distance)), alpha = 0.3,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() + scale_fill_viridis_d() +
  facet_wrap(~ deforestation_type, ncol = 4) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Max Habitat Patch Size") + ylab("Shannon Diversity") +
  theme(
    legend.position = c(0.9,0.8)
    )

#
# summarize data mean patch size
#
mdl1 <- mdl %>% mutate(habitat_patch_size = habitat_patch_size+1) %>%
                         group_by(birds_dispersal_distance, habitat_patch_size,birth_rate_birds,deforestation_type) %>% summarise(across(c(birds_at_deforestation:count_birds,mean_patch,degraded_mfp,non_degraded_mfp),list(mean=mean, sd=sd, ql=~quantile(.x,probs=c((1-0.95)/2)), qh=~quantile(.x,probs=c(1-(1-0.95)/2)) ))) #  %>% mutate(deltaPop = (count_birds - birds_at_deforestation )/ birds_at_deforestation  )
names(mdl1)

@@

mdl1 %>%  filter(birth_rate_birds == 2) %>% 
  ggplot(aes(x = mean_patch_mean, y = calc_number_of_species_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = .5) +
  geom_point(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_number_of_species_ql, ymax = calc_number_of_species_qh, group = interaction(birds_dispersal_distance, deforestation_type), fill = factor(birds_dispersal_distance)), alpha = 0.3,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() + scale_fill_viridis_d() +
  facet_wrap(~ deforestation_type, ncol = 2) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Mean Habitat Patch Size") + ylab("Number of species") +
  theme(
    legend.position = c(0.9,0.8)
    )

ggsave("Figures/AcrossLoss_Species_Dispersal_Lambda2_mean.png",width=8,height=6,units="in",dpi=600)

# Shannon-Diversity  
#

@@

ggsave("Figures/AcrossLoss_Shannon_Dispersal_Lambda2_mean.png",width=8,height=6,units="in",dpi=600)

### 2023-12-04 ms_SLOSS.Rmd
Updated manuscript

We considered only one amount of habitat destruction in combination with different community dynamics, neutral and competitive, taking into account healthy and mpoverished communities with different dispersal related traits, and different patterns of habitat loss.

@@

The parameters encompass a range of $growth-rates$ while holding the $death-rate$ fixed at 1. In this model, the relationship between these parameters, $\lambda = growth-rate / death-rate$, is the main determinant of the fate of an individual of a species. In similar non-spatial models, this relationship must exceed 1 for a species to persist, but in spatially explicit models the value needs to be higher, around 1.6. Thus we used a minimum $\lambda$ of 1.7, we also fixed the $migration-rate$ at 0.0001. We simulate each model variant using the parameters shown in Table 1. For the neutral classical version only, we also simulate random block habitat patterns.

@@

The Netlogo model source code is available at Github <https://github.com/lsaravia/DynamicForestExtinction> and the R source code to perform simulations and analysis is located at <https://github.com/EcoComplex/multiSLOSS>.

@@

In our neutral model simulations with a migration rate of \(0.0001\) and regular habitat patterns, species richness varied with habitat patch sizes, influenced by dispersal distance and the \(\lambda\) parameter (Figure 1). Species richness increased with \(\lambda\), with the highest richness at smaller patch sizes for higher \(\lambda\) values. Conversely, a \(\lambda\) of 3 or less showed reduced richness at the smallest patch sizes for lower dispersal distances. The interplay of dispersal distances, habitat sizes, and \(\lambda\) impacts community survival and species richness. This suggests a threshold patch size influenced by these 3 parameters. Extended dispersal distances enhance species survival in scenarios with low \(\lambda\) and smaller patches but reduce richness beyond a certain threshold. Similar patterns are evident for Shannon diversity (Figure 2).

@@

For the simulations made with random block habitat patterns, the patches do not have the same sizes but there is a distribution with a minimum size determined by the patch-size parameter and a bigger maximum size. Again there is a simimlar pattern with a maximum richness after the threshold for $\lambda$ values less than 3 and for values greater or equal the maximum is at the lower habitat patch size (Figure S1 & S2).

![Relationship between number of species and habitat patch size for different versions of the model (across panels) and varying dispersal distances. The neutral is the clasical neutral model, species are equivalent and have the same parameters; in the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant; the neutral-with-habitat selection model, is a neutral model but propagules search an empty patch so their survival is ensured. Simulations were performed using a fixed growth rate (2), migration and a replacement rate of 0.0001 and 0.3 respectively. Following 200 time steps of the model, 60% of the habitat was destroyed, rendering it unavailable for the species. The destruction of habitat resulted in the formation of regular patterns, where all remaining patches exhibited equal sizes.](Figures/AcrossBehav_Species_Dispersal_Lambda2_regular.png)

![Relationship between Shannon Diversity and habitat patch size for different versions of the model (across panels) and varying dispersal distances. The neutral is the clasical neutral model, species are equivalent and have the same parameters; in the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant; the neutral-with-habitat selection model, is a neutral model but propagules search an empty patch so their survival is ensured. Simulations were performed using a fixed growth rate (2), migration and a replacement rate of 0.0001 and 0.3 respectively. Following 200 time steps of the model, 60% of the habitat was destroyed, rendering it unavailable for the species. The destruction of habitat resulted in the formation of regular patterns, where all remaining patches exhibited equal sizes.](Figures/AcrossBehav_Shannon_Dispersal_Lambda2_regular.png)

@@

Our research contributes to the ongoing Single Large vs Several Small (SLOSS) debate in conservation biology, offering new insights based on our model findings. Consistent with previous studies, the results from our simulations primarily support the SS > SL pattern for species retention, as observed across various model scenarios [@Fahrig2020]. This trend underlines the greater efficacy of multiple smaller habitats in preserving species diversity compared to a single large habitat under similar conditions of habitat loss. A critical factor in this dynamic is the balance between reproduction and extinction rates ($\lambda$ parameter) relative to the size of the habitat patches, influencing their role as sources of propagules. Notably, the SL > SS scenario emerges only when dispersal limitations prevent patch connectivity, and patch sizes are below a certain threshold to prevent local extinctions, aligning with Rybicki's findings [@Rybicki2013].

In neutral models one species by chance tends to dominate a patch [@Hubbell2001], so in fragmented landscapes and once the species increments its density is very unprobable that another species replaces it so with more patches the probability of having more species is augmented. This effect is enhanced when individuals can select the habitat, thus fragmentation have a positive effect in the number of species. For competitive communities, there is a higher probability that different species dominate different smaller patches, compared with a patch. Thus, again fragmentation has a positive effect on number of species. 
This positive effect is counteracted by disperal, high dispersal rates tended to lead to a homogenization of species distributions, so fragmented landscapes tend to be more like not fragmented ones reducing overall species richness. This pattern was particularly pronounced for species combining high dispersal capabilities with competitive strengths, indicating a complex interplay between dispersal, competitive interactions, and species diversity [@Fahrig2020].

The spatial arrangement of habitat patches, contrary to some expectations, did not produce different species richness patterns regarding fragmentation with few exceptions. Instead, the interplay between dispersal, patch size and between patch distance emerged as the primary factor influencing biodiversity patterns following habitat fragmentation, corroborated by empirical studies in large carnivores [@Paviolo2016]. The exceptions could be important, because stressed communities---represented in our models with lowest $\lambdas$--- with low dispersal may have a continuous increment of richnes with patch size, observed only in random block degradation patterns. This could be related to the optimal patch size for random block degradations tends to be higher than for regular degradation patterns, so we could hypothesize that the optimal was outside the simulated range for these stressed communities.

Our findings support Fahrig's [-@Fahrig2021] conclusion that conditions favoring SL > SS are limited and infrequently observed in nature [@Fahrig2021]. Under conditions of adequate patch size and dispersal capabilities, the SS scenario typically retains more species. This underscores the ecological significance of smaller habitat patches, such as isolated trees, in maintaining biodiversity and landscape connectivity [@Shafer1995; @Manning2006]. Our results thus challenge the conventional preference for larger reserves in conservation planning and suggest a more nuanced, context-specific approach.

We also observed that the temporal dynamics of species extinction post-fragmentation vary, with long-term species loss in SS scenarios generally less severe than short-term extinctions in SL scenarios. This highlights the importance of considering both short-term and long-term impacts of habitat fragmentation on biodiversity.

@@

>>>>

We considered only one amount of habitat destruction in combination with different community dynamics, neutral and competitive, taking into account healthy and empoverished communities with different dispersal related traits, and different patterns of habitat loss.

@@

The parameters encompass a range of $growth-rates$ while holding the $death-rate$ fixed at 1. In this model, the relationship between these parameters, $\lambda = growth-rate / death-rate$, is the main determinant of the fate of an individual of a species. In similar non-spatial models, this relationship must exceed 1 for a species to persist, but in spatially explicit models the value needs to be higher, around 1.6. Thus we used a minimum $\lambda$ of 1.7, we also fixed the $migration-rate$ at 0.0001. We simulate each model variant using the parameters shown in Table 1. For the neutral classical version only, we also simulate random block habitat patterns. For each combination of parameters we run 30 simulations up to 1500 time steps and recorded the number of species, the Shannon Diversity index [@Magurran2013] and the total number of individuals. For some representative combiations of parameters we take snapshots of the model output showing the typical dynamics on time of the 3 recorded quantities and the rank abundance species distribution [@Newman2005; @May1975]

@@

The model was developed in Netlogo 6.3.0 and model source code is available at Github <https://github.com/lsaravia/DynamicForestExtinction>, the Overview, Design concepts, Details (ODD), protocol for describing individual-based models is include within the model source code [@Grimm2010]. We performed the simulations and analisis usin R version 4.3.1 [@RCoreTeam2023], the R source code to perform simulations and analysis is located at <https://github.com/EcoComplex/multiSLOSS>.

@@

In our neutral model simulations with a migration rate of \(0.0001\) and regular habitat patterns, species richness varied with habitat patch sizes, influenced by dispersal distance and the \(\lambda\) parameter (Figure 1). Species richness increased with \(\lambda\), with the highest richness at smaller patch sizes for higher \(\lambda\) values. Conversely, a \(\lambda\) of 3 or less showed reduced richness at the smallest patch sizes for lower dispersal distances. The interplay of dispersal distances, habitat sizes, and \(\lambda\) impacts community survival and species richness. For $\lambda < 4$ this suggests a threshold patch size influenced by these 3 parameters, for $\lambda = 4$ patch size decrease species richness. The differences between increasing $\lambda$'s almost dissapear when dispersal distance is higher. Extended dispersal distances could enhance species survival in scenarios with low \(\lambda\) and smaller patches but reduce richness beyond a certain threshold. Similar patterns are evident for Shannon diversity (Figure 2).

@@

For the simulations made with random block habitat patterns, the patches do not have the same sizes but there is a distribution with a minimum size determined by the patch-size parameter and a bigger maximum size (Figures S9 - S11). The pattern is similar to the regular case,  with a maximum richness for intermediate patch sizes and for values of $\lambda$ higher than 2 the maximum is at the lower habitat patch size (Figure S1 & S2). The sizes of patches for the random block simulations are bigger so species richness and diversity are lower, except for the minimum patch size (Figure S7 & S8). Examples of simulations with random block and regular habitat patterns are shown in the supplementary material (Figures S9 & S32). In these figures we can observe that for the minimum patch size only migrants survive for regular habitat patterns with neutral and hierarchical models.

![Relationship between number of species and habitat patch size for different versions of the model (across panels) and varying dispersal distances. The neutral is the clasical neutral model, species are equivalent and have the same parameters; in the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant; the neutral-with-habitat-selection model, is a neutral model but propagules search an empty patch so their survival is ensured. Simulations were performed using a fixed growth rate (2), migration and a replacement rate of 0.0001 and 0.3 respectively. Following 200 time steps of the model, 60% of the habitat was destroyed, rendering it unavailable for the species. The destruction of habitat resulted in the formation of regular patterns, where all remaining patches exhibited equal sizes.](Figures/AcrossBehav_Species_Dispersal_Lambda2_regular.png)

![Relationship between Shannon Diversity and habitat patch size for different versions of the model (across panels) and varying dispersal distances. The neutral is the clasical neutral model, species are equivalent and have the same parameters; in the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant; the neutral-with-habitat-selection model, is a neutral model but propagules search an empty patch so their survival is ensured. Simulations were performed using a fixed growth rate (2), migration and a replacement rate of 0.0001 and 0.3 respectively. Following 200 time steps of the model, 60% of the habitat was destroyed, rendering it unavailable for the species. The destruction of habitat resulted in the formation of regular patterns, where all remaining patches exhibited equal sizes.](Figures/AcrossBehav_Shannon_Dispersal_Lambda2_regular.png)

Comparing the three versions of the model at $\lambda =2$, for higher dispersal the neutral model tend to have higher richness and Shannon diversity at higher patch sizes, followed by the hierarchical model and the neutral-with-habitat-selection model (Figure 3 & 4). At the lowest patch size, only in the neutral-with-habitat-selection have populations that can survive, because in the other model versions the individuals observed are migrants that cannot establish a local populations an become extinct (See Figures S5 - S15). Thus an intermediate patch size with a maximum species richness is not observed for this model.

Examples of the final pattern of species and habitat distributions is shown in Figures S9- S32. Species spatial distributions closely match the spatial distribution previous to the habitat destruction for neutral and neutral-with-habitat-selection models. Is also clearly observed that for the hierarchical model there is a strong dominance of the more competitive species where Shannon Diversity is very low for higher dispersal rate (Figure 4 & S23).

@@

Our research contributes to the ongoing Single Large vs Several Small (SLOSS) debate in conservation biology, offering new insights based on our model findings. Consistent with previous studies, the results from our simulations primarily support the SS > SL pattern for species retention, as observed across various model scenarios [@Fahrig2020]. This trend underlines the greater efficacy of multiple smaller habitats in preserving species diversity compared to a single large habitat under similar conditions of habitat loss. A critical factor in this dynamic is the balance between reproduction and extinction rates ($\lambda$ parameter) and dispersal relative to the size of the habitat patches, influencing their role as sources of propagules. Notably, the SL > SS scenario emerges only when dispersal limitations prevent patch connectivity, and patch sizes are above a certain threshold to prevent local extinctions, aligning with Rybicki's [-@Rybicki2020] findings.

In neutral models one species by chance tends to dominate a patch [@Hubbell2001], so in fragmented landscapes and once the species increments its density is very unprobable that another species replaces it so with more patches the probability of having more species is augmented. This effect is enhanced when individuals can select the habitat, because of the increased survival of the propagules, thus fragmentation have a positive effect in the number of species. 
For competitive communities, there is a higher probability that different species dominate different smaller patches, compared with a bigger patch. Thus, again fragmentation has a positive effect on number of species, this effect was called stochastic divergence [@Riva2023]. In smaller patches there is an increase of extinction risk because they support smaller populations [@Laurance2002;@Shafer1995], however as patches become smaller they also become more numerous, reducing the probability of extinction over the entire set of patches due to spreading the risk effect [@Crowley1977]. The combination of these two effects, the stochastic divergence and the risk effect, explain the highest number of species with patch size over the threshold of species survival.

This positive effect is counteracted by dispersal, high dispersal rates tended to a homogenization of species distributions, so fragmented landscapes tend to be more like not fragmented ones reducing overall species richness. This pattern was particularly pronounced for species combining high dispersal capabilities with competitive strengths, indicating a complex interplay between dispersal, competitive interactions, and species diversity [@Fahrig2020].

Edge effects are present in the neutral and hierarchical models, but not in the neutral-with-habitat-selection model. In the first two models, the edge effect is due to the fact that individuals in the border of the patch have less neighbors and the probability of extinction is higher.  In the neutral-with-habitat-selection model, the edge effect is not present because the individuals search for empty patches, so the individuals in the border of the patch have the same probability of extinction that the individuals in the interior of the patch. Thus when patches are smaller the edge proportion is higher and the edge effect is more important, but this plays a more important role for the minimal patch size for survival of species than for the species richness.

The spatial arrangement of habitat patches, contrary to some expectations, did not produce different species richness patterns regarding habitat patch size. The exceptions are stressed communities---represented in our models with lowest $\lambda$--- with low dispersal may have a continuous increment of richness with patch size, observed only in random block degradation patterns. This could be related to the optimal patch size for random block habitat loss patterns tends to be higher than for regular patterns, so we could hypothesize that the optimal was outside the simulated range for these stressed communities. The aggregation of patches in the random block pattern (implies aggregation of the degraded habitat) could be positive if individual habitat patches are very small, because it allows some patches to surpass the minimum size required for survival. For bigger patch sizes it could be detrimental because the aggregation produces even bigger patch sizes that decrease species richness. A similar effect was observed on experimental landscapes [@Fletcher2023], aggregated patterns of habitat loss are detrimental for patch occupancy and eventually could produce the same effects observed in our models about richness.

Naturally, there are limitations to our modeling work. Firstly, it is important to note that more research is needed to apply our to determine the effect of fragmentation in real-world communities. We assume that all species have the same parameters, thus hypothetically determining these parameters for all the species, and using the mean to determine the responses to fragmentation would render an approximate value for the effects and thresholds. Also an improved version of the model with different values of the parameters for each species could be developed and compared with the neutral model. The effects of fragmentation could be influenced by many other ecological/behavioral characteristics of the species not included in the model, environmental heterogeneity, also the size of the focal landscape compared to dispersal distances, that determines the scale of the dynamics could influence quantitative results [@Rybicki2020]. Besides that, the approach of using mean values of the parameters of relatively simple multispecies models has been applied and tested to predict species biomass [@Fort2018].

In our simulations we fixed the amount of habitat loss to 60%, first it in previous simulation studies it was shown that  fragmentation was important when the habitat amount is low, around 10% [@Hanski2015; @Rybicki2013]. but it has been observed that the amount of habitat loss could influence the effect of fragmentation [@Loke2019; @With2016]. It was suggested that with very low habitat amount (e.g 7%) the arrangement of patches does not matters, this could paired with our results of stressed communities

Our findings support Fahrig's [-@Fahrig2021] conclusion that conditions favoring SL > SS are limited and infrequently observed in nature. Under conditions of adequate patch size and dispersal capabilities, the SS scenario typically retains more species. This underscores the ecological significance of smaller habitat patches, such as isolated trees, in maintaining biodiversity and landscape connectivity [@Yan2021; @Manning2006; @Shafer1995]. Our results thus challenge the conventional preference for larger reserves in conservation planning and support the paradigm shift in conservation towards considering sets of many small patches generally having  higher biodiversity than sets of few large patches [@Riva2023].

@@

<!-- We also observed without a systematic exploration the temporal dynamics of species extinction, short-term extinctions are more severe than long-term species loss (extinction debt), but extinctions can continue several years. The importance of long term species loss is lower in our model than in similar neutral models [@Thompson2019] This highlights the importance of considering both short-term and long-term impacts of habitat fragmentation on biodiversity. 
-->
